maze-dataset

### 2025-04-05 docs/paper/paper.tex
temp affiliation fix
https://github.com/understanding-search/maze-dataset/pull/55#discussion_r2026567100
changes from comments from CDB up to pg6, rebuild paper

\author[1%
%
\ensuremath\mathparagraph]{Michael Igorevich Ivanitskiy%
  \,\orcidlink{0000-0002-4213-4993}\,%
}
\author[%
%
]{Aaron Sandoval%
  \,\orcidlink{0009-0002-8380-6140}\,%
}
\author[2%
%
]{Alex F. Spies%
  \,\orcidlink{0000-0002-8708-1530}\,%
}
\author[%
%
]{Tilman Räuker%
  \,\orcidlink{0009-0009-6321-4413}\,%
}
\author[1%
%
]{Brandon Knutson%
  \,\orcidlink{0009-0004-8413-0239}\,%
}
\author[1%
%
]{Cecilia Diniz Behn%
  \,\orcidlink{0000-0002-8078-5105}\,%
}
\author[1%
%
]{Samy Wu Fung%
  \,\orcidlink{0000-0002-2926-4582}\,%
}

\affil[1]{Colorado School of Mines, Department of Applied Mathematics
and Statistics%
}
\affil[2]{Imperial College London%
}
\affil[$\mathparagraph$]{Corresponding author}
\date{\vspace{-2.5ex}}

@@

Solving mazes is a classic problem in computer science and artificial
intelligence, and humans have been constructing mazes for thousands of
years. Although finding the shortest path through a maze is a solved
problem, this very fact makes it an excellent testbed for studying how
machine learning algorithms solve problems and represent spatial
information. We introduce \texttt{maze-dataset}, a Python library for
generating, processing, and visualizing datasets of mazes. This library
supports a variety of maze generation algorithms providing mazes with or
without loops, fully-connected or not, and many other variations. These
generation algorithms can be configured with various parameters, and the
resulting mazes can be filtered to satisfy desired properties. Also
provided are tools for converting mazes to and from various formats
suitable for a variety of neural network architectures, such as
rasterized images, tokenized text sequences, and various visualizations.
As well as providing a simple interface for generating, storing, and
loading these datasets, \texttt{maze-dataset} is extensively tested,
type hinted, benchmarked, and documented.

@@

The generation of mazes with a given algorithm is not inherently a
complex task, but the ability to seamlessly switch out algorithms,
modify algorithm parameters, or filter by desired properties all while
preserving the ability to convert between different representations of
the maze is not trivial. This library aims to greatly streamline the
process of generating and working with datasets of mazes that can be
described as subgraphs of an \(n \times n\) lattice with boolean
connections and, optionally, start and end points that are nodes in the
graph. Furthermore, we place emphasis on a wide variety of possible text
output formats aimed at evaluating the spatial reasoning capabilities of
Large Language Models and other text-based transformer models.

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we have found
that existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) do not support flexible maze generation
algorithms that provide fine-grained control of generation parameters
and the ability to easily transform between multiple representations of
the mazes (Raster, Textual, Tokenized) for training and testing models.

@@

A multitude of public and open-source software packages exist for
generating mazes (\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). However, nearly all of these packages generate
and store mazes in a form that is not optimized for storage space or,
more importantly, computer readability. The mazes produces by other
packages are usually rasterized or in some form of image, rather than
the underlying graph structure, and this makes it difficult to work with
these datasets.

\begin{itemize}
\item
  Most prior works provide mazes in some kind of image or raster format,
  and we provide a variety of similar output formats:

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to that used in
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}), but provides over 5.8 million unique formats for
  converting mazes to a text stream, detailed in
  \autoref{tokenized-output-formats}.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. We use an extremely efficient method detailed in
  \autoref{implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

Furthermore, a dataset of mazes can be filtered to satisfy certain
properties. Custom filters can be specified, and some filters are
included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}.
For example, we can require a minimum path length from origin to target
of 3:

@@

All implemented maze generation algorithms are stochastic by nature. For
reproducibility, the \texttt{seed} parameter of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig}{\texttt{MazeDatasetConfig}}
may be set. In practice, we do not find that exact duplicates of mazes
are generated with any meaningful frequency, even when generating large
datasets.

@@

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information, and a connection list optimized
for storing sub-graphs of a lattice. These objects can be converted to
and from several formats to maximize their utility in different
contexts.

@@

\begin{figure}
\hypertarget{fig:e2h-raster}{%
\centering
\includegraphics[width=0.3\textwidth,height=\textheight]{figures/maze-raster-input-target.pdf}
\caption{Input is the rasterized maze without the path marked (left),
and provide as a target the maze with all but the correct path removed.
Configuration options exist to adjust whether endpoints are included and
if empty cells should be filled in.}\label{fig:e2h-raster}
}
\end{figure}

@@

We convert mazes to token sequences in two steps. First, the maze is
stringified using
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{as\_tokens()}}.
The
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class provides a powerful interface for configuring maze stringification
behavior. Second, the sequence of strings is tokenized into integers
using \texttt{encode()}. Tokenization uses a fixed vocabulary for
simplicity. Mazes up to \(50 \times 50\) are supported using unique
tokens, and up to \(128 \times 128\) when using coordinate tuple tokens.

There are many algorithms by which one might tokenize a 2D maze into a
1D format usable by autoregressive text models. Training multiple models
on the encodings output from each of these algorithms may produce very
different internal representations, learned solution algorithms, and
levels of performance. To allow exploration of how different maze
tokenization algorithms affect these models, the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class contains a rich set of options to customize how mazes are
stringified. This class contains 19 discrete parameters, resulting in
over 5.8 million unique tokenizers. But wait, there's more! There are 6
additional parameters available in the library which are untested but
further expand the the number of tokenizers by a factor of \(44/3\) to
86 million.

All output sequences consist of four token regions representing
different features of the maze of which we see an example in
\autoref{fig:token-regions}.

\begin{figure} 
  \centering
  \begin{minipage}{5in}
    \footnotesize
    \input{figures/outputs-tokens-colored.tex}
  \end{minipage}
  \caption{
    Example text output format with token regions highlighted.
    \colorbox[RGB]{ 217,210,233 }{Adjacency list}: text representation of the lattice graph,
    \colorbox[RGB]{ 217,234,211 }{Origin}: starting coordinate,
    \colorbox[RGB]{ 234,209,220 }{Target}: ending coordinate,
    \colorbox[RGB]{ 207,226,243 }{Path}: maze solution sequence
  }
  \label{fig:token-regions}
\end{figure}

@@

The tokenizer architecture is purposefully designed such that adding and
testing a wide variety of new tokenization algorithms is fast and
minimizes disturbances to functioning code. This is enabled by the
modular architecture and the automatic inclusion of any new tokenizers
in integration tests. To create a new tokenizer, developers forking the
library may simply create their own
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclass and implement the abstract methods. If the behavior change is
sufficiently small, simply adding a parameter to an existing
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclass and updating its implementation will suffice. For small
additions, simply adding new cases to existing unit tests will suffice.

@@

\begin{figure}
\hypertarget{fig:benchmarks}{%
\centering
\includegraphics[width=0.95\textwidth,height=\textheight]{figures/benchmarks/gridsize-vs-gentime.pdf}
\caption{Plots of maze generation time. Generation time scales
exponentially with maze size for all algorithms (left). Generation time
does not depend on the number of mazes being generated, and there is
minimal overhead to initializing the generation process for a small
dataset (right). Wilson's algorithm is notably less efficient than
others and has high variance. Note that for both plots, values are
averaged across all parameter sets for that algorithm, and
parallelization is disabled.}\label{fig:benchmarks}
}
\end{figure}

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), we allow placing additional constraints in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}},
such as enforcing that the start or end point be in a ``dead end'' with
only one accessible neighbor cell. However, combining this with cyclic
mazes (such as those generated with percolation), as was required for
the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. However, our
package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

@@

The base function learned by symbolic regression is not particularly
insightful, and is potentially subject to change. It is defined as
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#cfg_success_predict_fn}{\texttt{cfg\_success\_predict\_fn}},
and takes a 5 dimensional float vector created by
\texttt{MazeDatasetConfig.\_to\_ps\_array()} which represents the 0)
percolation value 1) grid size 2) endpoint deadend configuration 3)
endpoint uniqueness 4) categorical generation function index.

@@

\begin{itemize}
\item
  (\protect\hyperlink{ref-nolte2024multistep}{Nolte et al., 2024}) use
  \texttt{maze-dataset} to compare the effectiveness of transformers
  trained with the MLM-\(\mathcal{U}\)
  (\protect\hyperlink{ref-MLMU-kitouni2024factorization}{Kitouni et al.,
  2024}) multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  (\protect\hyperlink{ref-wang2024imperative}{Wang et al., 2024}) and
  (\protect\hyperlink{ref-chen2024iaimperative}{Chen et al., 2024}) use
  \texttt{maze-dataset} to study the effectiveness of imperative
  learning.
\end{itemize}

@@

This work was partially funded by National Science Foundation awards
DMS-2110745 and DMS-2309810. We are also grateful to LTFF and FAR Labs
for hosting three of the authors for a Residency Visit, and to various
members of FAR's technical staff for their advice.

@@

@@

>>>>

\author[1,3%
%
\ensuremath\mathparagraph]{Michael Igorevich Ivanitskiy%
  \,\orcidlink{0000-0002-4213-4993}\,%
}
\author[2%
%
]{Aaron Sandoval%
  \,\orcidlink{0009-0002-8380-6140}\,%
}
\author[2,3%
%
]{Alex F. Spies%
  \,\orcidlink{0000-0002-8708-1530}\,%
}
\author[2%
%
]{Tilman Räuker%
  \,\orcidlink{0009-0009-6321-4413}\,%
}
\author[1%
%
]{Brandon Knutson%
  \,\orcidlink{0009-0004-8413-0239}\,%
}
\author[1%
%
]{Cecilia Diniz Behn%
  \,\orcidlink{0000-0002-8078-5105}\,%
}
\author[1%
%
]{Samy Wu Fung%
  \,\orcidlink{0000-0002-2926-4582}\,%
}

\affil[1]{Colorado School of Mines, Department of Applied Mathematics
and Statistics%
}
\affil[2]{Imperial College London%
}
\affil[3]{unsearch.org%
}
\affil[4]{Independent%
}
\affil[$\mathparagraph$]{Corresponding author}
\date{\vspace{-2.5ex}}

@@

Solving mazes is a classic problem in computer science and artificial
intelligence, and humans have been constructing mazes for thousands of
years. Although finding the shortest path through a maze is a solved
problem, this very fact makes it an excellent testbed for studying how
machine learning algorithms solve problems and represent spatial
information. We introduce \texttt{maze-dataset}, a user-friendly Python
library for generating, processing, and visualizing datasets of mazes.
This library supports a variety of maze generation algorithms providing
mazes with or without loops, mazes that are connected or not, and many
other variations. These generation algorithms can be configured with
various parameters, and the resulting mazes can be filtered to satisfy
desired properties. Also provided are tools for converting mazes to and
from various formats suitable for a variety of neural network
architectures, such as rasterized images, tokenized text sequences, and
various visualizations. As well as providing a simple interface for
generating, storing, and loading these datasets, \texttt{maze-dataset}
is extensively tested, type hinted, benchmarked, and documented.

@@

While maze generation itself is straightforward, the architectural
challenge comes from building a system supporting many algorithms with
configurable parameters, property filtering, and representation
transformation. This library aims to greatly streamline the process of
generating and working with datasets of mazes that can be described as
subgraphs of an \(n \times n\) lattice with boolean connections and,
optionally, start and end points that are nodes in the graph.
Furthermore, we place emphasis on a wide variety of possible text output
formats aimed at evaluating the spatial reasoning capabilities of Large
Language Models and other text-based transformer models.

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we found that
existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) lack support for transforming between multiple
representations and provide limited control over the maze generation
process.

@@

A multitude of public and open-source software packages exist for
generating mazes (\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). However, nearly all of these packages produce
mazes represented as rasterized images or other visual formats rather
than the underlying graph structure, and this makes it difficult to work
with these datasets.

\begin{itemize}
\item
  Most prior works provide mazes in visual or raster formats, and we
  provide a variety of similar output formats:

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to that used in
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}), but provides over 5.8 million unique formats for
  converting mazes to a text stream, detailed in
  \autoref{tokenized-output-formats}.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. We use a highly efficient method detailed in
  \autoref{implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

Furthermore, a dataset of mazes can be filtered to satisfy certain
properties. Custom filters can be specified, and some filters are
included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}.
For example, we can require a minimum path length of three steps from
the origin to the target:

@@

All implemented maze generation algorithms are stochastic by nature. For
reproducibility, the \texttt{seed} parameter of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig}{\texttt{MazeDatasetConfig}}
may be set. In practice, using provided deduplication filters, we do
find that exact duplicate mazes are generated very infrequently, even
when generating very large datasets.

@@

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information and a tensor optimized for storing
sub-graphs of a lattice. These objects can be converted to and from
several formats to maximize their utility in different contexts.

@@

\begin{figure}
\hypertarget{fig:e2h-raster}{%
\centering
\includegraphics[width=0.3\textwidth,height=\textheight]{figures/maze-raster-input-target.pdf}
\caption{Input is the rasterized maze without the path marked (left),
and provide as a target the maze with all but the correct path removed
(right). Configuration options exist to adjust whether endpoints are
included and if empty cells should be filled in.}\label{fig:e2h-raster}
}
\end{figure}

@@

We convert mazes to token sequences in two steps. First, the maze is
stringified using
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{as\_tokens()}}.
The
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class provides a powerful interface for configuring maze stringification
behavior. Second, the sequence of strings is tokenized into integers
using \texttt{encode()}. Tokenization uses a fixed vocabulary for
simplicity. Mazes up to \(50 \times 50\) are supported when using a
unique token for each position, and up to \(128 \times 128\) are
supported when positions in the maze are represented as a pair of
coordinates.

There are many algorithms by which one might tokenize a 2D maze into a
1D format usable by autoregressive text models. Training multiple models
on the encodings output from each of these algorithms may produce very
different internal representations, learned solution algorithms, and
levels of performance. To allow exploration of how different maze
tokenization algorithms affect these models, the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class contains a rich set of options to customize how mazes are
stringified. This class contains 19 discrete parameters, resulting in
over 5.8 million unique tokenizers. There are 6 additional parameters
available whose functionality is not verified via automated testing, but
further expand the the number of tokenizers by a factor of \(44/3\) to
86 million.

All output sequences consist of four token regions representing
different features of the maze; an example output sequence is shown in
\autoref{fig:token-regions}.

\begin{figure} 
  \centering
  \begin{minipage}{5in}
    \footnotesize
    \input{figures/outputs-tokens-colored.tex}
  \end{minipage}
  \caption{
    Example text output format with token regions highlighted.
    \colorbox[RGB]{ 217,210,233 }{Adjacency list}: text representation of the graph,
    \colorbox[RGB]{ 217,234,211 }{Origin}: starting coordinate,
    \colorbox[RGB]{ 234,209,220 }{Target}: ending coordinate,
    \colorbox[RGB]{ 207,226,243 }{Path}: maze solution sequence
  }
  \label{fig:token-regions}
\end{figure}

@@

The tokenizer architecture is purposefully designed such that adding and
testing a wide variety of new tokenization algorithms is fast and
minimizes disturbances to functioning code. This is enabled by the
modular architecture and the automatic inclusion of any new tokenizers
in integration tests. To create a new variety of tokenizer, developers
forking the library may simply create their own
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclass and implement the abstract methods. If the behavior change is
sufficiently small, simply adding a parameter to an existing
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#_TokenizerElement}{\texttt{\_TokenizerElement}}
subclass and updating its implementation will suffice.

@@

\begin{figure}
\hypertarget{fig:benchmarks}{%
\centering
\includegraphics[width=0.95\textwidth,height=\textheight]{figures/benchmarks/gridsize-vs-gentime.pdf}
\caption{Plot of maze generation time. Generation time scales
exponentially with maze size for all algorithms. Generation time per
maze does not depend on the number of mazes being generated, and there
is minimal overhead to initializing the generation process for a small
dataset. Wilson's algorithm is notably less efficient than others and
has high variance. Note that values are averaged across all parameter
sets for that algorithm.}\label{fig:benchmarks}
}
\end{figure}

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), we allow for additional constraints in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}},
such as enforcing that the start or end point be in a ``dead end'' with
only one accessible neighbor cell. However, combining this with cyclic
mazes (such as those generated with percolation), as was required for
the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. However, our
package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

@@

The base function learned by symbolic regression privdes limited insight
and may be subject to change. It is defined as
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#cfg_success_predict_fn}{\texttt{cfg\_success\_predict\_fn}},
and takes a 5 dimensional float vector created by
\texttt{MazeDatasetConfig.\_to\_ps\_array()} which represents the 0)
percolation value 1) grid size 2) endpoint deadend configuration 3)
endpoint uniqueness 4) categorical generation function index.

@@

\begin{itemize}
\item
  By (\protect\hyperlink{ref-nolte2024multistep}{Nolte et al., 2024}) to
  compare the effectiveness of transformers trained with the
  MLM-\(\mathcal{U}\)
  (\protect\hyperlink{ref-MLMU-kitouni2024factorization}{Kitouni et al.,
  2024}) multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By (\protect\hyperlink{ref-wang2024imperative}{Wang et al., 2024}) and
  (\protect\hyperlink{ref-chen2024iaimperative}{Chen et al., 2024}) to
  study the effectiveness of imperative learning.
\item
  By (\protect\hyperlink{ref-zhang2025tscend}{Zhang et al., 2025}) to
  introduce a novel framework for reasoning diffusion models.
\item
  By (\protect\hyperlink{ref-dao2025alphamaze}{Dao \& Vu, 2025}) to
  improve spatial reasoning in LLMs with GRPO.
\end{itemize}

@@

This work was partially funded by National Science Foundation awards
DMS-2110745 and DMS-2309810. We are also grateful to LTFF and FAR Labs
for hosting MII, AFS, and TR for a residency visit, and to various
members of FAR's technical staff for their advice.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dao2025alphamaze}{}}%
Dao, A., \& Vu, D. B. (2025). AlphaMaze: Enhancing large language
models' spatial intelligence via GRPO. \emph{arXiv Preprint
arXiv:2502.14669}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-zhang2025tscend}{}}%
Zhang, T., Pan, J.-S., Feng, R., \& Wu, T. (2025). T-SCEND: Test-time
scalable MCTS-enhanced diffusion model. \emph{arXiv Preprint
arXiv:2502.01989}.
